---
title: "PE manuscript stats"
author: "Mirvie"
date: "2024-12-30"
output: html_document
---

This script is for reproducing the IN-TEXT statistical claims in the manuscript. Figures and tables are delegated to auxiliary scripts, as well as the GSEA/GO analysis and CV notebook.


```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

Load data
```{r}
df <- arrow::read_feather("sample_data.feather")  %>% 
  dplyr::rename(is_plac_hdp = is_pappa2_hdp) # renamed during a revision
gene_spaces <- jsonlite::read_json("genes_space.json")
all_genes <- unlist(gene_spaces$all_genes)
```


# Abstract

Correlation between PAPPA2 and delivery GA.
```{r}
# Implementation details:
# - Pearson correlation
# - Filtered to pe only
cor_pappa2_dga <- df %>%
  select(PAPPA2, is_pe, delivery_ga) %>%
  filter(
    is_pe == TRUE
  ) %>%
  correlation::cor_test(data = ., x = "PAPPA2", y = "delivery_ga", method = "pearson")
print(cor_pappa2_dga)
print(cor_pappa2_dga$p)
print(psychometric::CI.Rsq(rsq = cor_pappa2_dga$r**2, k = 1, n = cor_pappa2_dga$n_Obs))
```

Cohen's d of PAPPA2 for PAPPA2+ HDP vs. CTRL within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Filtered to NMRF samples only
# - Comparison is plac PE vs everything else.
reference_group <- df %>%
  select(is_plac_hdp, is_nmrf, PAPPA2) %>%
  filter(
    is_plac_hdp == FALSE,
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(is_plac_hdp, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(is_plac_hdp) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

# Introduction

## Study overview

Sample size of training.
```{r}
nrow(df)
```
## Molecular subtypes reclassify common clinical phenotypes

Cohen's d of PAPPA2 for PE over all training.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - All PE vs Not PE
effect_pappa2_pe <- effectsize::cohens_d(df$PAPPA2[df$is_pe == TRUE],
  y = df$PAPPA2[df$is_pe == FALSE],
  adjust = FALSE,
  pooled_sd = TRUE
)
print(effect_pappa2_pe)
print("")
print(effect_pappa2_pe$Cohens_d)
```

Cohen's d of CD163 for PE over all training.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - All PE vs Not PE
effect_cd163_pe <- effectsize::cohens_d(df[df$is_pe == TRUE, "CD163"]$CD163,
  y = df[df$is_pe == FALSE, "CD163"]$CD163,
  adjust = FALSE,
  pooled_sd = TRUE
)
print(effect_cd163_pe)
print("")
print(effect_cd163_pe$Cohens_d)
```

PAPPA2 Cohen's d for Plac PE and other HDP vs healthy controls
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - "joint Cohen's d" is interpreted to mean giving all the relevant samples the same label and then computing effect size
# - Reference group is hdp_severity == "control" (i.e. no PE, no GHTN, no ambiguity in the label)
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(hdp_severity == "control") %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, hdp_vanity_group, is_nmrf, PAPPA2) %>%
  filter(hdp_vanity_group != "PE, Ambiguous") %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

```{r}
# Calculate uncertainty for the fold-change difference in effect sizes using bootstrapping.
# Includes p-value and confidence interval. Note that p-value should be expressed as an inequality relative
# to the number of bootstrap iters.
ctrl_vector <- reference_group
case_vector <- df[(df$hdp_vanity_group != "PE, Ambiguous") & (df$hdp_severity == "preterm_pe"), "PAPPA2" ]$PAPPA2
other_vector <- df[(df$hdp_vanity_group != "PE, Ambiguous") & (df$hdp_severity == "other_hdp"), "PAPPA2" ]$PAPPA2
print(effectsize::cohens_d(y = ctrl_vector,
                       x = case_vector)$Cohens_d / 
        effectsize::cohens_d(y = ctrl_vector,
                       x = other_vector)$Cohens_d)
boot_iters <- 50000
case_results <- array(dim = boot_iters)
other_results <- array(dim = boot_iters)
for (boot_idx in 1:boot_iters){
  ctrl_sample <- sample(ctrl_vector, replace = TRUE)
  case_results[boot_idx] <- effectsize::cohens_d(y = ctrl_sample,
                       x = sample(case_vector, replace = TRUE))$Cohens_d
  other_results[boot_idx] <- effectsize::cohens_d(y = ctrl_sample,
                       x = sample(other_vector, replace = TRUE))$Cohens_d
}
ratios <- case_results / other_results
print(quantile(ratios, c(0.025, 0.975)))
print(mean(ratios < 1))

```

CD163 Cohen's d for Plac PE and other HDP vs healthy controls
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - "joint Cohen's d" is interpreted to mean giving all the relevant samples the same label and then computing effect size
# - Reference group is hdp_severity == "control" (i.e. no PE, no GHTN, no ambiguity in the label)
reference_group <- df %>%
  select(hdp_severity, is_nmrf, CD163) %>%
  filter(hdp_severity == "control") %>%
  pull(CD163)
effect_sizes <- df %>%
  select(hdp_severity, hdp_vanity_group, is_nmrf, CD163) %>%
  filter(hdp_vanity_group != "PE, Ambiguous") %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(CD163, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```


NMRF and MRF sample size in training.
```{r}
df %>%
  select(is_nmrf) %>%
  group_by(is_nmrf) %>%
  summarise(n = n())
```

Cohen's d of PAPPA2 for PAPPA2+ PE and Other HDP vs healthy controls within MRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == False
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == FALSE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == FALSE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

Cohen's d of PAPPA2 for PAPPA2+ PE and Other HDP vs healthy controls within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == True
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```


Cohen's d of PAPPA2 for PAPPA2+ PE vs. all other samples within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity %in% c("control", "other_hdp")
# - Filtered to NMRF == True
# - Use the hdp_severity label to identify controls, to align with prior analyses.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(
    hdp_severity %in% c("control", "other_hdp"),
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(is_plac_hdp, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(is_plac_hdp) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```


Correlation between PAPPA2 and delivery GA within NMRF, by PAPPA2+ status.
```{r}
# Implementation details:
# - Pearson correlation
# - Filtered to NMRF samples only
# - Filtered to green triangle only
cor_pappa2_dga <- df %>%
  select(PAPPA2, is_nmrf, is_plac_hdp, delivery_ga) %>%
  filter(
    is_nmrf == TRUE,
    is_plac_hdp == TRUE
  ) %>%
  correlation::cor_test(data = ., x = "PAPPA2", y = "delivery_ga", method = "pearson")
print(cor_pappa2_dga)
print(cor_pappa2_dga$p)
print(psychometric::CI.Rsq(rsq = cor_pappa2_dga$r**2, k = 1, n = cor_pappa2_dga$n_Obs))
```

Cohen's d of PAPPA2 for non-PE spontaneous preterm birth
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Move PE to control (i.e. these are sPTB "due to preeclampsia")
# - Postpartum PE still included in the sPTB.
df$sptb_not_pe <- FALSE
df[(df$is_pe == FALSE) & (df$is_sptb == TRUE), "sptb_not_pe"] <- TRUE
print(sum(df$sptb_not_pe))
print(effectsize::cohens_d(df$PAPPA2[df$sptb_not_pe == TRUE], 
                           y = df$PAPPA2[df$sptb_not_pe == FALSE], 
                           adjust = FALSE, pooled_sd = TRUE))
```

```{r}
sessionInfo()
```
