---
title: "PE manuscript stats"
author: "Mirvie"
date: "2024-06-29"
output: html_document
---

This script is for reproducing the IN-TEXT statistical claims in the manuscript. Figures and tables are delegated to auxiliary scripts, as well as the GSEA/GO analysis and CV notebook.


```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

Load data
```{r}
df <- arrow::read_feather("sample_data.feather")
gene_spaces <- jsonlite::read_json("genes_space.json")
all_genes <- unlist(gene_spaces$all_genes)
```


# Abstract

Correlation between PAPPA2 and delivery GA.
```{r}
# Implementation details:
# - Pearson correlation
# - Filtered to NMRF samples only
# - Filtered to green triangle only
cor_pappa2_dga <- df %>%
  select(PAPPA2, is_nmrf, is_pappa2_hdp, delivery_ga) %>%
  filter(
    is_nmrf == TRUE,
    is_pappa2_hdp == TRUE
  ) %>%
  correlation::cor_test(data = ., x = "PAPPA2", y = "delivery_ga", method = "pearson")
print(cor_pappa2_dga)
print(cor_pappa2_dga$p)
print(psychometric::CI.Rsq(rsq = cor_pappa2_dga$r**2, k = 1, n = cor_pappa2_dga$n_Obs))
```

Cohen's d of PAPPA2 for PAPPA2+ HDP vs. CTRL within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Filtered to NMRF samples only
# - Comparison is GT vs everything else.
reference_group <- df %>%
  select(is_pappa2_hdp, is_nmrf, PAPPA2) %>%
  filter(
    is_pappa2_hdp == FALSE,
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(is_pappa2_hdp, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(is_pappa2_hdp) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

# Introduction

## Study overview

Sample size of training.
```{r}
nrow(df)
```
## Molecular phenotypes of disease

Cohen's d of PAPPA2 for PE over all training.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - All PE vs Not PE
effect_pappa2_pe <- effectsize::cohens_d(df$PAPPA2[df$is_pe == TRUE],
  y = df$PAPPA2[df$is_pe == FALSE],
  adjust = FALSE,
  pooled_sd = TRUE
)
print(effect_pappa2_pe)
print("")
print(effect_pappa2_pe$Cohens_d)
```

Cohen's d of CD163 for PE over all training.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - All PE vs Not PE
effect_cd163_pe <- effectsize::cohens_d(df[df$is_pe == TRUE, "CD163"]$CD163,
  y = df[df$is_pe == FALSE, "CD163"]$CD163,
  adjust = FALSE,
  pooled_sd = TRUE
)
print(effect_cd163_pe)
print("")
print(effect_cd163_pe$Cohens_d)
```

Cohen's d for PAPPA2+ PE and other HDP vs healthy controls
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - "joint Cohen's d" is interpreted to mean giving all the relevant samples the same label and then computing effect size
# - Reference group is hdp_severity == "control" (i.e. no PE, no GHTN, no ambiguity in the label)
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(hdp_severity == "control") %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, hdp_vanity_group, is_nmrf, PAPPA2) %>%
  filter(hdp_vanity_group != "PE, Ambiguous") %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

NMRF and MRF sample size in training.
```{r}
df %>%
  select(is_nmrf) %>%
  group_by(is_nmrf) %>%
  summarise(n = n())
```

Cohen's d of PAPPA2 for PAPPA2+ PE and Other HDP vs healthy controls within MRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == False
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == FALSE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == FALSE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

Cohen's d of PAPPA2 for PAPPA2+ PE and Other HDP vs healthy controls within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == True
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA2)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PAPPA2) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA2, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  ) %>%
  arrange(desc(Cohens_d))
effect_sizes
```

Correlation between PAPPA2 and delivery GA within NMRF, by PAPPA2+ status.
```{r}
# Implementation details:
# - Pearson correlation
# - Filtered to NMRF samples only
# - Filtered to green triangle only
cor_pappa2_dga <- df %>%
  select(PAPPA2, is_nmrf, is_pappa2_hdp, delivery_ga) %>%
  filter(
    is_nmrf == TRUE,
    is_pappa2_hdp == TRUE
  ) %>%
  correlation::cor_test(data = ., x = "PAPPA2", y = "delivery_ga", method = "pearson")
print(cor_pappa2_dga)
print(cor_pappa2_dga$p)
print(psychometric::CI.Rsq(rsq = cor_pappa2_dga$r**2, k = 1, n = cor_pappa2_dga$n_Obs))
```

Cohen's d of PAPPA2 for non-PE spontaneous preterm birth
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Move PE to control (i.e. these are sPTB "due to preeclampsia")
# - Postpartum PE still included in the sPTB.
df$sptb_not_pe <- FALSE
df[(df$is_pe == FALSE) & (df$is_sptb == TRUE), "sptb_not_pe"] <- TRUE
print(sum(df$sptb_not_pe))
print(effectsize::cohens_d(df$PAPPA2[df$sptb_not_pe == TRUE], 
                           y = df$PAPPA2[df$sptb_not_pe == FALSE], 
                           adjust = FALSE, pooled_sd = TRUE))
```

## RNA-driven Classifier to Predict PAPPA2+ HDP

Cohen's d of PAPPA (not PAPPA2) for PAPPA2+ PE and Other HDP vs. healthy controls within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == True
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PAPPA) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == TRUE
  ) %>%
  pull(PAPPA)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PAPPA) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PAPPA, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  )
effect_sizes
```

Cohen's d of PLGF for PAPPA2+ PE and Other HDP vs. healthy controls within NMRF.
```{r}
# Implementation details:
# - Cohen's d with no small-sample adjustment and a pooled SD
# - Reference group is hdp_severity == "control" (i.e. exclude GHTN, non-GT PE)
# - Filtered to NMRF == True
# - Use the hdp_severity label to identify GT instead of the GT label, to align with identification of controls.
reference_group <- df %>%
  select(hdp_severity, is_nmrf, PGF) %>%
  filter(
    hdp_severity == "control",
    is_nmrf == TRUE
  ) %>%
  pull(PGF)
effect_sizes <- df %>%
  select(hdp_severity, is_nmrf, PGF) %>%
  filter(is_nmrf == TRUE) %>%
  group_by(hdp_severity) %>%
  summarise(effectsize::cohens_d(PGF, y = reference_group, adjust = FALSE, pooled_sd = TRUE),
    n = n()
  )
effect_sizes
```

```{r}
sessionInfo()
```
